---
title: "ENVX1002_Group_Project"
format: 
  html:
    embed-resources: true
    code-background: true

---

# **-NOT THE FINAL DOCUMENT-**
I'll make a proper, professional looking one for final submission. This file is the raw data analysis. 

Intructions: Make a folder in your docs and a new project in RStudio. Create a sub-folder labelled "data" and copy the "WHR_2023" File from the World Happiness Index dataset into the new "data" folder. Open the relevant RStudio project and simply copy-paste all the code into a new quarto document to view everything below in the Quarto doc format that we all know and "love". 

# Outline

**EDA**

-   What are the variables like.

-   Central value and spread of variables.

-   Distribution of data.

**Modelling**

1.  Null and alternate hypotheses

2.  Fit the model

3.  check Assumptions

4.  interpret model output. State the equation, state estimates and what they mean.

5.  Interpret significance of model and predictors.

6.  Interpret model fit statistics. Residual standard error, appropriate R2, plot regression line.

7.  Statistical conclusion

8.  Scientific conclusion.

## Checklist

-   [x] EDA - central tendency stats
-   [x] EDA - Distribution stats
-   [x] EDA - QQ plots and histograms
-   [ ] EDA - log transform data - Currently stuck here.
-   [ ] Model - hypotheses, null and alternate
-   [ ] Model - fit the model
-   [ ] Model - Check assumptions
-   [ ] Model - output equation
-   [ ] Model - estimates and what they mean
-   [ ] Model - interpret significance of model and predictors
-   [ ] Model - Model fit statistics, R2, Residual SE, plot regression line
-   [ ] Model - statistical and scientific conclusions.

# Now here's the actually important stuff

Set up stuff:

```{r}
library(readr)
library(tidyverse)
library(modeest)
library(ggplot2)
```

```{r}
WHR_2023 <- read_csv("data/WHR_2023.csv")
gdp <- WHR_2023$gdp_per_capita
life <- WHR_2023$healthy_life_expectancy
happy <- WHR_2023$happiness_score
```

## EDA Section

Variable Description: Both GDP per capita and Healthy Life Expectancy are continuous, numerical variables.

There are no missing data points in GDP per capita. There is one missing data point in Healthy Life Expectancy. The missing data point is from the State of Palestine.

### Central tendency stats

**GDP:** Mean, median, mode.

```{r}
mean_gdp <- mean(gdp)
median_gdp <- median(gdp)
mode_gdp <- mfv(gdp)[1]
mean_gdp
median_gdp
mode_gdp
```

Summary: Mean = 1.406985. Median = 1.449. Mode = 0.914.

**Healthy life expectancy:** Mean, median, mode.

```{r}
mean_life <- mean(life, na.rm = TRUE)
median_life <- median(life, na.rm = TRUE)
mode_life <- mfv(life, na_rm = TRUE)[1]
mean_life
median_life
mode_life
```

Summary: Mean = 0.3661765. median = 0.3895, mode = 0.389.

**GDP:** Measures of Spread

```{r}
range_gdp <- range(gdp)
iqr_gdp <- IQR(gdp)
var_gdp <- var(gdp)
sd_gdp <- sd(gdp)
range_gdp
iqr_gdp
var_gdp
sd_gdp
```

Summary: Range = 0 to 2.2. Interquartile range (IQR) = 0.699. Variance = 0.1874567. Standard Deviation (SD) = 0.4329627.

```{r}
# Standard error = population sd / sqrt(n) = 0.43 / sqrt(137)
se_gdp <- 0.43/sqrt(137)
se_gdp
```
Summary: Standard Error of GDP per capita = 0.03673738

**Life Expectancy** Measures of Spread

```{r}
range_life <- range(life, na.rm = TRUE)
iqr_life <- IQR(life, na.rm = TRUE)
var_life <- var(life, na.rm = TRUE)
sd_life <- sd(life, na.rm = TRUE)
range_life
iqr_life
var_life
sd_life
```
Summary: Range = 0 to 0.702. Interquartile range (IQR) = 0.239. Variance = 0.0245521. Standard Deviation (SD) = 0.1566911.

```{r}
#Standard error = population sd / sqrt(n) = 0.16 / sqrt(136)
se_life <- 0.16/ sqrt(136)
se_life
```
Summary: Standard error of Life Expectancy = 0.01371989.

### Distribution Analysis

**GDP**
```{r}
ggplot(WHR_2023, aes(x = gdp))+
  geom_histogram(binwidth = 0.1, fill = "lightgreen", colour = "black")+
  xlab("GDP per capita")
```
First impressions: Data is not normal and is skewed to the right. [self-note] use log10 transformation

```{r}
ggplot(WHR_2023, aes(sample= gdp))+
  stat_qq()+
  stat_qq_line()+
  labs(title = "QQ plot of GDP per capita", x = "GDP per capita", y = "Count")
```
Statistical analysis of distribution:
```{r}
library(moments)
skewness(gdp)
kurtosis(gdp)
```
Summary: Skewness = -0.4782739. Kurtosis = 2.64509.

```{r}
shapiro.test(gdp)
```
Summary: The formal normality test indicates that the null hypothesis of the data following a normal distribution should be rejected (p-value = p-value = 0.001494). 
Therefore we can conclude that the data is positively skewed.


**Life Expectancy**
```{r}
ggplot(WHR_2023, aes(x = life))+
  geom_histogram(binwidth = 0.1, fill = "lightgreen", colour = "black")+
  xlab("Healthy Life Expectancy")
```
```{r}
ggplot(WHR_2023, aes(sample= life))+
  stat_qq()+
  stat_qq_line()+
  labs(title = "QQ plot of Healthy Life Expectancy", x = "Healthy Life Expectancy", y = "Count")
```
Statistical Analysis of distribution data:
```{r}
skewness(life, na.rm = TRUE)
kurtosis(life, na.rm = TRUE)
```
Summary: Skewness = -0.4038421. Kurtosis = 2.20732.

```{r}
shapiro.test(life)
```
The formal normality test indicates that the null hypothesis of the data following a normal distribution should be rejected (p-value = p-value = 0.0002804). 
Therefore we can conclude that the data is positively skewed.

```


### Transformations

**Transformation Codes:**
```{r}
#gdp
WHR_2023$log10_gdp <- log10(WHR_2023$gdp_per_capita) #new colum for log10 data
log10_gdp <- log10(gdp) #log10 transformation
natlog_gdp <- log1p(gdp) #natural log transformation
sqrt_gdp <- sqrt(gdp) #square root transformation
cbrt_gdp <- sign(gdp) * abs(gdp) ^ (1/3) #cube root transformation
```

```{r}
#life expectancy
WHR_2023$log10_life <- log10(WHR_2023$healthy_life_expectancy)
log10_life <- WHR_2023$log10_life
natlog_life <- log1p(life)
sqrt_life <- sqrt(life)
cbrt_life <- sign(life) * abs(life) ^ (1/3)
```

**Re-checking assumptions:**
```{r}
ggplot(WHR_2023, aes(x = sqrt_life))+
  geom_histogram(binwidth = 0.1, fill = "lightgreen", colour = "black")+
  xlab("GDP per capita")
```

**Formal Testing:**
```{r}
shapiro.test(cbrt_life)
```
*Problem solving notes for skewed GDP data:*
Log10 transformation: failed (dammit). Data still majorly skewed.
Shapiro-test result: NA

Natural log transformation: failed (double dammit). Data still skewed, less than log10 transformation.
Shaprio-test result: 1.302e-06. *Violates normality.*

Square root transformation: The least skewed in appearance.
Shapiro-test result: 1.194e-08. *Violates normality.*

Cube root transformation: As bad as the log10 transformation.
Shaprio-test result: 5.168e-13. *Violates normality*


*Problem Solving for skewed Life Expectancy data:*
Log10 Transformation: failed. Shapiro-test result = NA.

Natural Log transformation: failed. Shapiro-test result = 1.79e-05.*Violates normality*

Square root transformation: failed. shapiro-test result = 2.477e-07.*Violates normality*

Cube root transformation: failed. Shapiro-test result = 6.272e-10.*Violates normality*
