---
title: "ENVX1002_Group_Project"
format: 
  html:
    embed-resources: true
    code-background: true

---
# Data Analysis Report for ENVX1002 Group 3

Set up stuff:

```{r}
library(readr)
library(tidyverse)
library(modeest)
library(ggplot2)
```

```{r}
WHR_2023 <- read_csv("data/WHR_2023.csv")
gdp <- WHR_2023$gdp_per_capita
life <- WHR_2023$healthy_life_expectancy
happy <- WHR_2023$happiness_score
```

## EDA Section

Variable Description: Both GDP per capita and Healthy Life Expectancy are continuous, numerical variables.

There are no missing data points in GDP per capita. There is one missing data point in Healthy Life Expectancy. The missing data point is from the State of Palestine.

### Central tendency stats

**GDP:** Mean, median, mode.

```{r}
mean_gdp <- mean(gdp)
median_gdp <- median(gdp)
mode_gdp <- mfv(gdp)[1]
mean_gdp
median_gdp
mode_gdp
```

Summary: Mean = 1.406985. Median = 1.449. Mode = 0.914.

**Healthy life expectancy:** Mean, median, mode.

```{r}
mean_life <- mean(life, na.rm = TRUE)
median_life <- median(life, na.rm = TRUE)
mode_life <- mfv(life, na_rm = TRUE)[1]
mean_life
median_life
mode_life
```

Summary: Mean = 0.3661765. median = 0.3895, mode = 0.389.

**GDP:** Measures of Spread

```{r}
range_gdp <- range(gdp)
iqr_gdp <- IQR(gdp)
var_gdp <- var(gdp)
sd_gdp <- sd(gdp)
range_gdp
iqr_gdp
var_gdp
sd_gdp
```

Summary: Range = 0 to 2.2. Interquartile range (IQR) = 0.699. Variance = 0.1874567. Standard Deviation (SD) = 0.4329627.

```{r}
# Standard error = population sd / sqrt(n) = 0.43 / sqrt(137)
se_gdp <- 0.43/sqrt(137)
se_gdp
```

Summary: Standard Error of GDP per capita = 0.03673738

**Life Expectancy:** Measures of Spread

```{r}
range_life <- range(life, na.rm = TRUE)
iqr_life <- IQR(life, na.rm = TRUE)
var_life <- var(life, na.rm = TRUE)
sd_life <- sd(life, na.rm = TRUE)
range_life
iqr_life
var_life
sd_life
```

Summary: Range = 0 to 0.702. Interquartile range (IQR) = 0.239. Variance = 0.0245521. Standard Deviation (SD) = 0.1566911.

```{r}
#Standard error = population sd / sqrt(n) = 0.16 / sqrt(136)
se_life <- 0.16/ sqrt(136)
se_life
```

Summary: Standard error of Life Expectancy = 0.01371989.

### Distribution Analysis

**GDP**

```{r}
ggplot(WHR_2023, aes(x = gdp))+
  geom_histogram(binwidth = 0.1, fill = "lightgreen", colour = "black")+
  xlab("GDP per capita")
```

First impressions: Data is not normal and is skewed to the right. 

```{r}
ggplot(WHR_2023, aes(sample= gdp))+
  stat_qq()+
  stat_qq_line()+
  labs(title = "QQ plot of GDP per capita", x = "GDP per capita", y = "Count")
```

**Statistical analysis of distribution:**

```{r}
library(moments)
skewness(gdp)
kurtosis(gdp)
```

Summary: Skewness = -0.4782739. Kurtosis = 2.64509.

```{r}
shapiro.test(gdp)
```

Summary: The formal normality test indicates that the null hypothesis of the data following a normal distribution should be rejected (p-value = p-value = 0.001494). Therefore we can conclude that the data is positively skewed.

**Life Expectancy**

```{r}
ggplot(WHR_2023, aes(x = life))+
  geom_histogram(binwidth = 0.1, fill = "lightgreen", colour = "black")+
  xlab("Healthy Life Expectancy")
```

```{r}
ggplot(WHR_2023, aes(sample= life))+
  stat_qq()+
  stat_qq_line()+
  labs(title = "QQ plot of Healthy Life Expectancy", x = "Healthy Life Expectancy", y = "Count")
```

**Statistical Analysis of distribution data:**

```{r}
skewness(life, na.rm = TRUE)
kurtosis(life, na.rm = TRUE)
```

Summary: Skewness = -0.4038421. Kurtosis = 2.20732.

```{r}
shapiro.test(life)
```

The formal normality test indicates that the null hypothesis of the data following a normal distribution should be rejected (p-value = p-value = 0.0002804). Therefore we can conclude that the data is positively skewed.

# Regression Model Section
Scatterplot
```{r}
plot(gdp, life)
plot(life, gdp)
```

**Correlation coefficient:**
```{r}
cor(gdp, life, use = "complete.obs")
```
Output: 0.8375183


## 1. Hypothesis

Aim: To determine whether the amount of GDP per capita (GDP) increases Healthy Life Expectancy (Life)

H0: Slope of the fitted model = 0, therefore GDP per capita is not a significant predictor of Healthy Life Expectancy, and the mean of Healthy Life Expectancy is a better estimate. We could not conclude that GDP per capita does not adequately explain the variation in Healthy Life Expectancy.

H1: Slope of the fitted model =/= 0, therefore we can say that GDP per capita is a significant predictor of Healthy Life Expectancy.


## 2. Fit the model
```{r}
model.lm <- lm(life ~ gdp, data = WHR_2023)
```

## 3. Checking Assumptions
```{r}
par(mfrow = c(2,2))
plot(model.lm)
```
Data appears to meet all assumptions, therefore the model fits and thus data analysis can proceed without transformations.

```{r}
summary(model.lm)
```
**Output Summary:**

Coefficients: Intercept Estimate = - 0.05988. Life expectancy estimate = 0.30240.

Residual standard error: 0.08594 on 134 DF.

Multiple R-squared: 0.7014. Adjusted R-squared: 0.6992. F-statistic: 314.8 on 1 and 134 DF. P-value <2.2e-16.

## 4. Regression Equation and estimates

**Regression Line equation:**

Basic equation: intercept = y-mean - gradient * x-mean

GDP = -0.06 - 0.30 * Healthy Life expectancy

## Adding the linear model to the scatterplot.
```{r}
plot(life, gdp, xlab = "GDP per capita", ylab = "Healthy Life Expectancy")
abline(model.lm, col = "red")
```

## 5. Conclusions

**Statistical Conclusion:**
As the p value (< 2.2e-16) is below 0.05, the null hypothesis is rejected. 

There is a significant relationship between healthy life expectancy and GDP per capita (linear regression: R^2 = 0.70, F = 314.80, df = 1, 134, p<2.2e-16). 

Healthy Life Expectancy = 0.302 x GDP per capita - 0.060

**Biological Conclusion:**

GDP per capita is a significant predictor of healthy life expectancy (p = < 2.2e-16) - it accounts for 70% of the variation found in healthy life expectancy. 
